Fusing Filters with Integer Linear Programming

### 2014-04-12 2014betterfusionforfilters/section/01-Introduction.tex
Add notes and move sections around

\item
We present a simplification to constraint generation that is also applicable to some existing integer linear programming formulations such as Megiddo's,
where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two. \TODO{ref}

\item
Our constraint system also encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program.
For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced.
\TODO{ref}

>>>>

\item
We present a simplification to constraint generation that is also applicable to some existing integer linear programming formulations such as Megiddo's,
where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two (\S\ref{s:OptimisedConstraints}).

\item
Our constraint system also encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program. For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced (\S\ref{s:ObjectiveFunction}).

### 2014-04-12 2014betterfusionforfilters/section/03-SizeInference.tex
More work on intro, suggested title change
Add notes and move sections around

\subsection{Runtime checks on array lengths}
The $@map@_n$ combinator requires all its input arrays to be the same size.
We do not, however, want to perform runtime checks on the size of arrays before every combinator.
We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.
Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.

We use the constraints to rule out any programs that require runtime checks on array lengths during the program.
Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.
Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid.
Finally, any initial equality constraints, such as two input arrays being the same length, or one input array being the length of the cross product of two others, are dealt with by inserting runtime checks only at the start of the function.

>>>>

\subsection{Runtime checks on array lengths}
The $@map@_n$ combinator requires all its input arrays to be the same size.
We do not, however, want to perform runtime checks on the size of arrays before every combinator.
We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.

\TODO{We assign different rates to both @ys1@ and @ys2@ in this example.}
\begin{code}
 diff xs = let ys1 = filter p xs
               ys2 = filter p xs
           in  (ys1, ys2)
\end{code}

% \ben{This point is only relavant in the context of our specific implementation, which is no the focus of this paper. }
% Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.

We use the constraints to rule out any programs that require runtime checks on array lengths during the program. Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.

\ben{Refactor presentation to use the actual $\exists k. $ quantifier, as in ``The essence of ML type inference'', from ATAPL}

Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid. Finally, any initial equality constraints, such as two input arrays being the same length, or one input array being the length of the cross product of two others, are dealt with by inserting runtime checks only at the start of the function.

### 2014-04-12 2014betterfusionforfilters/section/04-ILP.tex
More work on intro, suggested title change
Add notes and move sections around

%!TEX root = ../Main.tex
\section{Integer linear program formulation}

@@

@@

\subsubsection{The simple version}
\paragraph{Acyclic and precedence-preserving}

@@

If there is no edge between $i$ and $j$, the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively means there is no constraint on the $\pi$ values.
Note, however, that other edges may still constrain $i$ or $j$ by transitivity.

@@

\paragraph{Fusion between different types}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators.
\TODO{Expand}.

@@

\subsubsection{The objective function}

To find the objective function, note that fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.

@@

\subsubsection{An optimised version}
We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
Likewise with clusters of different, incompatible types.
The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.

>>>>

%!TEX root = ../Main.tex
\section{Integer linear program formulation}
\label{s:ILP}

@@

\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }

@@

\subsubsection{The simple version}
\TODO{Why is this ``the simple version''? Why present a simple version before the real one?} \TODO{We need to show a complete constraint system at the start. Maybe use a simple objective function to start with, then refine it in a later section.}

\paragraph{Acyclic and precedence-preserving}

@@

If there is no edge between $i$ and $j$, the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively means there is no constraint on the $\pi$ values. Note, however, that other edges may still constrain $i$ or $j$ by transitivity. \ben{If there is ``no constraint'' on the $\pi$ values then why bother adding the constraint at all?}

@@

\paragraph{Fusion between different types}

\ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators. \ben{What is a subrate? How do we determine the subrates of each rate?}
\TODO{Expand}.

@@

\subsubsection{The objective function}
\label{s:ObjectiveFunction}

To find the objective function, note that fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion. \TODO{Refer to example code that demonstrates these different opportunities.}

@@

\subsubsection{An optimised version}
\label{s:OptimisedConstraints}
We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
Likewise with clusters of different, incompatible types.
The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.
